A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-28 arxiv/proof_main.tex

\begin{proof}[Proof of Theorem \ref{thm_many_tasks}]
	We first introduce several notations.
	Let $\cal W = [W_1, W_2, \dots, W_t] \in \real^{r\times t}$ be a matrix that contains all the parameters of the output layers.
	Let $U_X U_X^{\top} = X (X^{\top} X)^{-1} X^{\top} \in\real^{n \times n}$ denote the projection matrix to $X$.

	Recall that all tasks have the same covariates in this setting.
	Therefore, the optimization objective equation \eqref{eq_mtl} becomes the following
	\begin{align}
		f(B; W_1, \dots, W_t) = \sum_{i=1}^t \bignorm{X B W_i - Y_i}^2, \label{eq_mtl_same_cov}
	\end{align}
	where $B \in \real^{p \times r}$ and $W_1, W_2, \dots, W_t \in \R^r$ for $1 < r < t$ by our assumption.
	Using the local optimality condition over $B$, that is $\nabla_B f = 0$, we obtain a closed form solution $\hat{B}$ that depends on the output layers as follows
	\begin{align*}
		\hat{B}(W_1, \dots, W_t) &= (X^{\top}X)^{-1} X^{\top} \bigbrace{\sum_{i=1}^t Y_i W_i^{\top}} (\cal W  \cal W^{\top})^{-1} \\
		&= (B^\star \cal W ^{\top}) (\cal W \cal W ^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cal W \cal W^{\top})^{-1}.
	\end{align*}
	Plugging in the equation $\hat{B}$ back into equation \eqref{eq_mtl_same_cov} we obatin
	\begin{align}
		g(\cal W) = \sum_{j=1}^t \bignorm{X B^{\star} \cW^{\top} (\cW\cW^{\top})^{-1} W_j + U_X U_X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cW\cW^{\top})^{-1} W_j - Y_i}^2. \label{eq_mtl_output_layer}
	\end{align}
	Recall that we obtain $\hat{\cW}$ by minimizing $\cW$ in the above equation.
	In order to derive the minimizer, our proof involves two steps.
	First, we consider the expectation of $g(\cW)$ over $\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t$, and $X$.
	We show that the minimizer of $\ex{g(\cW)}$ has a simple closed form solution similar to principal component analysis.
	Second, we show that the concentration error between $g(\cW)$ and $\ex{g(\cW)}$ is small provided with $n$ samples.
	Intuitively, this is because $\cW$ only has $r \times t$ parameters, which is much smaller than $n$ by our assumption.
	Hence we use standard concentration bounds and $\epsilon$-net arguments to show that the concentration error is small for $g(\cW)$.
	To facilitate the analysis, we divide $g(\cW)$ into three parts based on their degree of dependence on the random noise:
	\be\nonumber
	g(\cal W)= g_0 + 2g_1 + g_2, 
	\ee
	where
	\begin{align*}
		g_0(\cal W) &= \sum_{j=1}^t \bignorm{XB^{\star} \cW^{\top} (\cW\cW^{\top})^{-1} W_j - X\beta_j}^2 = \bignorm{X B^{\star} (\cW^{\top} (\cW\cW^{\top})^{-1} \cW - \id_{t\times t})}^2, \\
		g_1(\cal W)  &= \sum_{j=1}^t \inner{XB^{\star} \cW^{\top} (\cW\cW^{\top})^{-1}W_j - X\beta_j}{U_X U_X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cW\cW^{\top})^{-1} W_j - \varepsilon_j} ,\\
				%&= \sum_{j=1}^t \inner{-X B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1} W_j + X\beta_j}{\varepsilon_j}, \\
		g_2(\cal W)  &= \sum_{j=1}^t \bignorm{U_X U_X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cW\cW^{\top})^{-1}W_j - \varepsilon_j}^2 .
					\end{align*}

>>>>

\begin{proof}[Proof of Theorem \ref{thm_many_tasks}]
	We first introduce several notations.
	Let $U_X U_X^{\top} = X (X^{\top} X)^{-1} X^{\top} \in\real^{n \times n}$ denote the projection matrix to $X$.

	Recall that we obtain $\hat{\cW}$ by minimizing $\cW$ in the above equation.
	In order to derive the minimizer, our proof involves two steps.
	First, we consider the expectation of $g(\cW)$ over $\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t$, and $X$.
	We show that the minimizer of $\ex{g(\cW)}$ has a simple closed form solution similar to principal component analysis.
	Second, we show that the concentration error between $g(\cW)$ and $\ex{g(\cW)}$ is small provided with $n$ samples.
	Intuitively, this is because $\cW$ only has $r \times t$ parameters, which is much smaller than $n$ by our assumption.
	Hence we use standard concentration bounds and $\epsilon$-net arguments to show that the concentration error is small for $g(\cW)$.
	To facilitate the analysis, we divide $g(\cW)$ into three parts based on their degree of dependence on the random noise:
	\be\nonumber
	g(\cal W)= g_0 + 2g_1 + g_2, 
	\ee
	where
	\begin{align*}
		g_0(\cal W) &= \sum_{j=1}^t \bignorm{XB^{\star} \cW^{\top} (\cW\cW^{\top})^{-1} W_j - X\beta_j}^2 = \bignorm{X B^{\star} (\cW^{\top} (\cW\cW^{\top})^{-1} \cW - \id_{t\times t})}^2, \\
		g_1(\cal W)  &= \sum_{j=1}^t \inner{XB^{\star} \cW^{\top} (\cW\cW^{\top})^{-1}W_j - X\beta_j}{U_X U_X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cW\cW^{\top})^{-1} W_j - \varepsilon_j} ,\\
				%&= \sum_{j=1}^t \inner{-X B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1} W_j + X\beta_j}{\varepsilon_j}, \\
		g_2(\cal W)  &= \sum_{j=1}^t \bignorm{U_X U_X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cW\cW^{\top})^{-1}W_j - \varepsilon_j}^2 .
					\end{align*}

### 2020-08-28 arxiv/setup.tex
reorganize

\noindent\textbf{Data model.}
Suppose we have $t$ labeled regression tasks. %where $t$ is a fixed value that does not grow with the feature dimension $p$.
For each task $k$ from $1$ to $t$, we have $n_k$ samples $x_1, x_2, \dots, x_{n_k}$ that are all $p$-dimensional feature vectors with real-valued labels $y_1, y_2, \dots, y_{n_k}$.
Without loss of generality, we focus on predicting the $t$-th task and refer the $t$-th task as the target.
We refer to task $1$ to task $t-1$ as source tasks.
We assume that the labels of each task $k$ satisfy the linear model with unknown parameters $\beta_k\in\real^p$, that is,
	\[ y_i = x_i^{\top}\beta_k + e_i, \text{ for all } 1 \le i \le n_k, \]
where $e_i$ denotes random noise with mean zero and variance $\sigma^2$.
To write the above notations more succiently, let $X_k \in \real^{n_k \times p}$ denote the covariates that consists of a feature vector in every row.
Let
	\[ Y_k = X_k \beta_k + \varepsilon_k, \text{ for all } 1\le i \le t \] denote the label vector of task $k$ and $\varepsilon_k$ denote an i.i.d. random vector with mean zero and variance $\sigma^2$.

@@

We make several standard assumptions on the covariance matrix $\Sigma_k$ and the random vector $z_i$, for $k = 1,\dots, t$ and $i = 1,\dots, n_k$.
First, for every entry of $z_i$, we assume that its fourth moment exists \todo{check}, that is, there exists a constant $C > 0$ such that
\begin{equation}\label{assmAhigh}
	\ex{\abs{z_{i,j}}^4} \le C,\ 1\le j \le p.
\end{equation}
Second, we assume that the spectral norm of $\Sigma_k$ is bounded by a fixed constant, for every $k = 1,\dots, t$.
This is without loss of generality since we can always rescale $\Sigma_k$ so that its spectral norm satisfies the condition.

>>>>

\noindent\textbf{Data model.}
Suppose we have $t$ labeled regression tasks. %where $t$ is a fixed value that does not grow with the feature dimension $p$.
For each task $k$ from $1$ to $t$, we have $n_k$ samples $x_1, x_2, \dots, x_{n_k}$ that are all $p$-dimensional feature vectors with real-valued labels $y_1, y_2, \dots, y_{n_k}$.
Without loss of generality, we focus on predicting the $t$-th task and refer the $t$-th task as the target.
We refer to task $1$ to task $t-1$ as source tasks.
We assume that the labels of each task $k$ satisfy the linear model with unknown parameters $\beta_k\in\real^p$, that is,
	\[ y_i = x_i^{\top}\beta_k + e_i, \text{ for all } 1 \le i \le n_k, \]
where $e_i$ denotes random noise with mean zero and variance $\sigma^2$.
To write the above notations more succiently, let $X_k \in \real^{n_k \times p}$ denote the covariates that consists of a feature vector in every row.
Let
	\[ Y_k = X_k \beta_k + \varepsilon_k, \text{ for all } 1\le k \le t \] denote the label vector of task $k$ and $\varepsilon_k$ denote an i.i.d. random vector with mean zero and variance $\sigma^2$.

@@

We make several standard assumptions on the covariance matrix $\Sigma_k$ and the random vector $z_i$, for $k = 1,\dots, t$ and $i = 1,\dots, n_k$.
First, for every entry of $z_i$, we assume that its $(4+\e)$-th moment exists, that is, there exist constants $\e, C > 0$ such that
\begin{equation}\label{assmAhigh}
	\ex{\abs{z_{i,j}}^{4+\e}} \le C,\ 1\le j \le p.
\end{equation}
Second, we assume that the spectral norm of $\Sigma_k$ is bounded by a fixed constant, for every $k = 1,\dots, t$.
This is without loss of generality since we can always rescale $\Sigma_k$ so that its spectral norm satisfies the condition.
Finally, for the noise vector $\varepsilon_k$, we assume that there exists an absolute constant such that for all $l$ \todo{check}, the $l$-th moment of the noise entries exists:
\begin{align*}
	\ex{\abs{\varepsilon_{k, j}}^l \le C}, \text{ for all } j = 1, \dots, n_k.
\end{align*}

### 2020-08-28 arxiv/transfer.tex

%\iffalse
Next, we derive a closed-form solution of the multi-task learning estimator for the case of two tasks.
From \cite{WZR20}, we know that we need to explicitly restrict the output dimension $r$ of $B$ so that there is transfer between the two tasks.
Hence for the case of two tasks, we consider the setting where $r=1$.
For simplicity of notations, we shall denote $(X_i^{tr},Y_i^{tr})$ and $(X_i^{val},Y_i^{val})$ as $(X_i,Y_i)$ and  $(\wt X_i,\wt Y_i)$, respectively. Then equation \eqref{eq_mtl} simplifies to
\begin{align}\label{eq_mtl_2task}
	f(B; w_1, w_2) = \bignorm{X_1 B w_1 - Y_1}^2 + \bignorm{X_2 B w_2 - Y_2}^2,
\end{align}
where $B\in\real^p$ and $w_1, w_2$ are both real numbers. To solve the above problem, suppose that $w_1, w_2$ are fixed, by local optimality, we find the optimal $B$ as
\begin{align}
	& \hat{B}(w_1, w_2) = (w_1^2 X_1^{\top}X_1 + w_2^2 X_2^{\top}X_2)^{-1} (w_1 X_1^{\top}Y_1 + w_2 X_2^{\top}Y_2) \label{hatB}\\
	&= \frac{1}{w_2} \left( \frac{w_1^2}{w_2^2}  X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1} \left(\frac{w_1}{w_2} X_1^{\top}Y_1 + X_2^{\top}Y_2\right) \nonumber\\
	&= \frac{1}{w_2}\left[\beta_2 + \left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}\bigbrace{X_1^{\top}X_1\left(\frac{w_1}{w_2}\beta_1 - \frac{w_1^2}{w_2^2} \beta_2\right) + \left(\frac{w_1}{w_2} X_1^{\top}\varepsilon_1 + X_2^{\top}\varepsilon_2\right)}\right]. \nonumber
\end{align}
As a remark, when $w_1 = w_2 = 1$, we obtain linear regression.
If $\beta_1$ is a scaling of $\beta_2$, then  $w_1, w_2$ can be scaled accordingly to fix both tasks more accurately than linear regression.

@@

Hence to minimize $g(v)$, it suffices to minimize $\val(v)$ over $v$.
Let $\hat v=\hat{w_1}/\hat{w_2}$ be the global minimizer of $g(v)$.
Now we can define the multi-task learning estimator for the target task as
	\[ \hat{\beta}_2^{\MTL} = \hat{w}_{2}\hat{B}(\hat{w}_1, \hat{w}_2) .\]
%	where $t=2$ since we are considering the two task case, and it also stands for the ``target task".
%The intuition for deriving $\hat{\beta}_2^{\MTL}$ is akin to performing multi-task training in practice.
%Let $\hat{v} = \hat{w_1} / \hat{w_2}$ for the simplicity of notation.
The prediction loss of using $\hat{\beta}_2^{\MTL}$ for the target task is
\begin{align}
	\te(\hat{\beta}_2^{\MTL}) =&~ \hat{v}^2 \bignorm{\Sigma_2^{1/2}(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1 (\beta_1 - \hat{v} \beta_2)}^2 \nonumber \\
			&+~  \bigtr{\Sigma_2(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-2}\left(\sigma_1^2 \cdot \hat v^2  X_1^{\top}X_1 + \sigma_2^2  \cdot X_2^{\top}X_2\right) }, \label{eq_te_mtl_2task}
\end{align}
which only depends on $\hat v$, the sample covariance matrices, and $\beta_1,\beta_2$.
%\fi

@@

Next, we describe our result for the multiple-task case where all tasks have the same covariates, that is, $X_i = X$ for all $i = 1, \dots, t$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
Similar to the two-task case, we consider an arbitrary local minimum $B, W_1, \dots, W_2$ of the optimization objective.
We extend the bias-variance decomposition from the two-task case to the multiple-task case.
More precisely, we observe that the expected prediction loss of $\hat{\beta}_t^{\MTL}$ conditional on $X$ consists of a bias and a variance equation as follows
\begin{align}
	\exarg{\varepsilon_1, \dots, \varepsilon_t}{L(\hat{\beta}_t^{\MTL}) \mid X}
	=& \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1} W_t - \beta_t}}^2 \label{eq_bias_multiple} \\
	&+ \sigma^2 \cdot (W_t^{\top} (\cW \cW^{\top})^{-1} W_t) \cdot \bigtr{\Sigma (X^{\top} X)^{-1}}
\end{align}

>>>>

\iffalse
Next, we derive a closed-form solution of the multi-task learning estimator for the case of two tasks.
From \cite{WZR20}, we know that we need to explicitly restrict the output dimension $r$ of $B$ so that there is transfer between the two tasks.
Hence for the case of two tasks, we consider the setting where $r=1$.
For simplicity of notations, we shall denote $(X_i^{tr},Y_i^{tr})$ and $(X_i^{val},Y_i^{val})$ as $(X_i,Y_i)$ and  $(\wt X_i,\wt Y_i)$, respectively. Then equation \eqref{eq_mtl} simplifies to
\begin{align}\label{eq_mtl_2task}
	f(B; w_1, w_2) = \bignorm{X_1 B w_1 - Y_1}^2 + \bignorm{X_2 B w_2 - Y_2}^2,
\end{align}
where $B\in\real^p$ and $w_1, w_2$ are both real numbers. To solve the above problem, suppose that $w_1, w_2$ are fixed, by local optimality, we find the optimal $B$ as
\begin{align}
	& \hat{B}(w_1, w_2) = (w_1^2 X_1^{\top}X_1 + w_2^2 X_2^{\top}X_2)^{-1} (w_1 X_1^{\top}Y_1 + w_2 X_2^{\top}Y_2) \label{hatB}\\
	&= \frac{1}{w_2} \left( \frac{w_1^2}{w_2^2}  X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1} \left(\frac{w_1}{w_2} X_1^{\top}Y_1 + X_2^{\top}Y_2\right) \nonumber\\
	&= \frac{1}{w_2}\left[\beta_2 + \left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}\bigbrace{X_1^{\top}X_1\left(\frac{w_1}{w_2}\beta_1 - \frac{w_1^2}{w_2^2} \beta_2\right) + \left(\frac{w_1}{w_2} X_1^{\top}\varepsilon_1 + X_2^{\top}\varepsilon_2\right)}\right]. \nonumber
\end{align}
As a remark, when $w_1 = w_2 = 1$, we obtain linear regression.
If $\beta_1$ is a scaling of $\beta_2$, then  $w_1, w_2$ can be scaled accordingly to fix both tasks more accurately than linear regression.

@@

Hence to minimize $g(v)$, it suffices to minimize $\val(v)$ over $v$.
Let $\hat v=\hat{w_1}/\hat{w_2}$ be the global minimizer of $g(v)$.
Now we can define the multi-task learning estimator for the target task as
	\[ \hat{\beta}_2^{\MTL} = \hat{w}_{2}\hat{B}(\hat{w}_1, \hat{w}_2) .\]
%	where $t=2$ since we are considering the two task case, and it also stands for the ``target task".
%The intuition for deriving $\hat{\beta}_2^{\MTL}$ is akin to performing multi-task training in practice.
%Let $\hat{v} = \hat{w_1} / \hat{w_2}$ for the simplicity of notation.
The prediction loss of using $\hat{\beta}_2^{\MTL}$ for the target task is
\begin{align}
	\te(\hat{\beta}_2^{\MTL}) =&~ \hat{v}^2 \bignorm{\Sigma_2^{1/2}(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1 (\beta_1 - \hat{v} \beta_2)}^2 \nonumber \\
			&+~  \bigtr{\Sigma_2(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-2}\left(\sigma_1^2 \cdot \hat v^2  X_1^{\top}X_1 + \sigma_2^2  \cdot X_2^{\top}X_2\right) }, \label{eq_te_mtl_2task}
\end{align}
which only depends on $\hat v$, the sample covariance matrices, and $\beta_1,\beta_2$.
\fi

@@

Next, we describe our result for the multiple-task case where all tasks have the same covariates, that is, $X_i = X$ and $n_i = n$ for all $i = 1, \dots, t$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
Similar to the two-task case, we consider an arbitrary local minimum $B, W_1, \dots, W_2$ of the optimization objective.
We extend the bias-variance decomposition from the two-task case to the multiple-task case.
Let $\cal W = [W_1, W_2, \dots, W_t] \in \real^{r\times t}$ be a matrix that contains all the parameters of the output layers.
We observe that the expected prediction loss of $\hat{\beta}_t^{\MTL}$ conditional on $X$ consists of a bias and a variance equation as follows
\begin{align}
	\exarg{\varepsilon_1, \dots, \varepsilon_t}{L(\hat{\beta}_t^{\MTL}) \mid X}
	=& \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1} W_t - \beta_t}}^2 \label{eq_bias_multiple} \\
	&+ \sigma^2 \cdot (W_t^{\top} (\cW \cW^{\top})^{-1} W_t) \cdot \bigtr{\Sigma (X^{\top} X)^{-1}} \label{eq_var_multiple}
\end{align}
One can see that equation \eqref{eq_bias_multiple} is the bias of the multi-task learning estimator and equation \eqref{eq_var_multiple} is its variance.
Compared to the prediction loss of single-task learning (cf. equation \eqref{eq_var_stl}), we observe that the variance equation \eqref{eq_var_multiple} is always smaller because $W_t^{\top} (\cW \cW^{\top})^{-1} W_t \le 1$.
On the other hand, the bias equation \eqref{eq_bias_multiple} is always larger because of the difference between the task models.

Our main result for the multiple-task case provides generalization bounds that relate the empirical bias and variance of multi-task learning to the sample size $n$ and the distance between task models.
